Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2018-05-09 WholePlotTrajectories.Rmd
Redundancy etc

>>>>

```{r RedFun, out.width = '100%', echo=FALSE, fig.cap="Trajectories of the functional redundancy over 30 years after disturbance. Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation and the missing trait value filling processes.Initial treatments are represented by solid lines colors with green for control, blue for T1,orange for T2 and red for T3.", fig.env="figure"}
load("DB/Redundancy")

RedundancyPlot(RedundancyTraj)
mtext("Functional Redundancy",side=3,adj=0,line=1.5)
mtext("Sum of abundance-weighted Trait Probability Density",side=3,adj=0,cex=0.8,line=0.5)
mtext("Years since disturbance",side=1,line=2.2,adj=1)

### 2018-05-09 WholePlotTrajectories.tex
Redundancy etc

\caption{\label{tab:Tab1}Intervention table, summary of the disturbance intensity for the 4 plot treatments in Paracou.}
\centering
\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
\toprule
Treatment & Timber & Thinning & Fuelwood & \%AGB lost\\
\midrule
Control &  &  &  & 0\\
T1 & DBH $\geq$ 50 cm, commercial species, $\approx$ 10 trees/ha &  &  & $[12\%-33\%]$\\
T2 & DBH $\geq$ 50 cm, commercial species, $\approx$ 10 trees/ha & DBH $\geq$ 40 cm, non-valuable species, $\approx$ 30 trees/ha &  & $[33\%-56\%]$\\
T3 & DBH $\geq$ 50 cm, commercial species, $\approx$ 10 trees/ha & DBH $\geq$ 50 cm, non-valuable species, $\approx$ 15 trees/ha & 40 cm $\leq$ DBH $\leq$ 50 cm, non-valuable species, $\approx$ 15 trees/ha & $[35\%-56\%]$\\
\bottomrule
\end{tabu}
\end{table}

@@

To counter the remaining taxonomic uncertainty plots taxonomic
composition and diversity were analysed at the genus level, \emph{i.e.}
referring to the genus of observed or trialed botanical names.
Variations in plots taxonomic and functional composition after
disturbance was visualized by their trajectories in a two-dimensional
ordination space over 30 years. Two NMDS were conducted, either from
taxonomic flora inventories or from plots functional composition based
on the 8 leaf, stem and life history traits. Plots trajectories along
time was reported comparatively to the inventories in 1989, 5 years
after disturbance, which corresponded to first inventory with a
sufficient degree of uncertainty (\textless{}30\% of undetermined
trees). The inventories dissimilarity compared to the reference 1989
inventory was reported using occurrence-based (Jaccard) and
abundance-based (Bray-Curtis) similarity measures. The trajectory of
inventories along time was visualized with the euclidean distance in the
two-dimensional ordination space to the 1989 inventory. The trajectories
of the leaf and stem and life traits were also visualized with the
community weighted means (CWM), representing the average trait value in
a community weighted by relative abundance of the species carrying each
value \citep{Diaz2007, Garnier2004}. To compensate the intrinsinc
difference among plots the trajectories corresponded to the differences
along time with the reference inventory in 1984 (5 year after
disturbance). Species seed mass corresponded to 5 classes of increasing
mass, seed mass trajectories were therefore reported as the proportion
of each class in the inventories.

@@

Over time, 828388 trees and 591 botanical species pertaining to 223
genus and 64 botanical families were recorded. Taxonomic and functional
plots trajectories were examined through the ordination in a two
dimensional ordiantion space of successive inventories from 1989 (5
years after disturbance) to 2015 (31 years after disturbance).
Classifications were performed using either abundance-based Bray-Curtis
(Figure \ref{fig:NMDSplans}) or incidence-based Jaccard dissimilarity,
this last having given similar results (data not shown).

@@

The changes observed in plots functional composition went hand to hand
with consistent trajectories of the 8 functional and life history traits
(Leaf thickness, chlorophyll content, toughness and specific area,wood
specific gravity and barck thickness and seed mass and maximum adult
height) visaualized with plots community weighted means (CWM) (Figure
\ref{fig:CWM}).

Except for leaf cholorphyll content, which continued to increase for
some T3 and T2 plots 30 years after disturbance, all traits and seed
mass proportions displayed a unimodal trajectories. The weighted means
of communities specific maximum height at adult stage (\emph{Hmax}),
leaf toughness (\emph{L\_toughness}) and wood specific gravity
(\emph{WD}) remained significantly lower than their initial value and
than these of the control plots (Figure \ref{fig:CWM}). The weighted
means of bark thickness (\emph{Bark\_thick}) similarly remained
substantilly higher than initially for all disturbed plots while the
specfic leaf area (\emph{SLA}) had almost recovered its initial value
and this of the undisturbed plots at the end of the experiment. Once
again the maximum difference to initial state was correlated to the
disturbance intensity (\(\rho_{spearman}^{L_thickness}=0.67\),
\(\rho_{spearman}^{L_chloro}=0.45\),
\(\rho_{spearman}^{L_toughness}=-0.43\), \(\rho_{spearman}^{SLA}=0.93\),
\(\rho_{spearman}^{WD}=-0.78\),
\(\rho_{spearman}^{Bark-thickness}=0.88\),
\(\rho_{spearman}^{Hmax}=-0.48\)).

@@

\caption{Trajectories of the communities weighted means (CWM) over 30 years after disturbance of 4 leaf traits (Leaf thickness, \emph{L\_thickness}, chlorophyll content, \emph{L\_chloro}, toughness, \emph{L\_toughness} and specific area, \emph{SLA}), 2 stem traits (wood specific gravity, \emph{WD}, and bark thickness, \emph{Bark-thick}) and one life trait (Specific maximum height at adult stage, \emph{Hmax}). Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation and the missing trait value filling processes. Initial treatments are represented by solid lines colorswith green for control, blue for T1,orange for T2 and red for T3.}\label{fig:CWM}
\end{figure*}

@@

For undisturbed plots the Richness, Shannon and Simpson diversity
remained stable over the 30 monitored years. After disturbance the
richness increased when the disturbance intensity was low, with a
maximum increase of 14 botanical genuses (plot 3 from treatment 2). When
the disturbance was intense however, plot staxonomic richness followed a
unimodal decrease with a return to initial values after intense
disturbance. The taxonomic evenness (Shannon and Simpson diversities)
significantly increased whenever the disturbance intensity. The evenness
followed a unimodal trajectory with a just beginning return towards
initial values and a maximum reached around 20 years and positively
correlated to the disturbance intensity
(\(\rho_{spearman}^{Shannon}=0.86\), and
\(\rho_{spearman}^{Simpson}=0.89\)). Only two T3 plots, plots 8 and 12,
remained increasing 30 years after disturbance \ref{fig:DivTaxo}),
suggesting a similar be much delayed unimodal trajectory.

@@

The functional diversity trajectories of plots based on the 8 leaf, stem
and life history traits was examined through the Rao diversity. For all
undisturbed plots the functional diversity remained comparable to the
reference values 5 years after disturbance (1989 inventories). For all
disturbed plots the functional trajectories followed a unimodal
trajectory with a maximum positively correlated to disturbance intensity
(\(\rho_{spearman}=0.73\)) (Figure \ref{fig:DivFun}). The plot 7 from
treatment 1 that displayed a constantly outlying diversity was removed
from the graphical representation for better readability, see appendix
for full graph. Thirty years after disturbance all plots, whenever the
initial disturbance intensity, regained diversity values similar to
their initial avule and to those of control plots.

@@

{\centering \includegraphics[width=1\linewidth]{WholePlotTrajectories_files/figure-latex/DivFun-1}

@@

\caption{Trajectories of the Rao functional diversity over 30 years after disturbance. Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation. Initial treatments are represented by solid lines colorswith green for control, blue for T1,orange for T2 and red for T3.and the missing trait value filling processes.}\label{fig:DivFun}
\end{figure}

@@

The monitoring of disturbed forest communities confirmed a limited
negative impact of disturbance on species richness, as observed on
several post logging surveys \citep{Cannon1998, Baraloto2012a}. Thirty
years after disturbance, the genus richness was restored to initial plot
values after high disturbance intensity while it substantially increased
after low disturbance intensity, reaching a gain of almost 12 genuses
for some plots.

Both richness and evenness after disturbance followed asymptotic
trajectories, sharply increasing for 15 years before stabilizing at
higher values than those of control plots. An increase of communities
evenness stems from a higher homogeneity of species distribution. After
disturbance, communities are made of the old, pre-disturbance survivors
and the newly recruited trees: changes in composition and abundance
distribution are to be found in the recruitment processes and in the
pre-disturbance survivors mortality. The composition of old survivors
proved to mirror the initial community \citep{Herault2018}, which leads
to support a species turnover among recruited trees with enhanced growth
and survival of previously infrequent species. Indeed, the increase in
taxonomic diversity was accompanied by an increase of taxonomic
dissimilarity with plots initial state and a functional shift towards
resource-acquisitive strategies (sharp increase in the SLA, leaf
thickness and bark thickness and decrease in wood density, leaf
toughness and maximum height)
\citep{Westoby1998, Wright2004, Reich2014}. Disturbance then causes a
reorganization of the typical high dominance structure of hyperdiverse
mature forests after disturbance, benefiting to pioneers and light
demanding species. Likely, the changes in abiotic environment and
competitive pressure favored pioneers which outcompete others in non
limiting resources but are excluded in mature forests by long-lived,
more resistant and shade tolerant species.

As stated by the IDH, communities dynamics after disturbance then rely
on species functional strategy and corresponding ability to fill the
environmental niches made available by disturbance. Recruited species
then mix with pre-disturbance ones from which they differ and constitue
a community all the more diversified that the disturbance is intense
\citep{Molino2001}.

@@

The trajectories of almost all leaf and stem functional traits, life
history traits,and functional diversity followed a hump-back trajectory
which maximum were positively correlated to the disturbance intensity,
and were recovering towards initial values after 30 years that
trajectory. Functoinal traits and communities functional diversity are
the most direct link between biodiversity and ecosystem functioning
\citep{Diaz2005}: these trajectories reflected a consistent recovery in
the long term of ecosystem processes \citep{Guariguata2001}.

Same hump-back trajectories were followed by the composition of plots
species and functoinal composition. This similarly meant a recovery of
the pre-disturbance state and the maintenance of the initial differences
among plots which supports the idea of a convergence of communities
composition \citep{Hubbell1999, Molino2001, Baraloto2012a}. Although
plots followed homogeneous trends these initial differences in
compositon difference, and probably also some abiotic parameters,
strongly determined the maximum and time path of the trajectories
\citep{Anderson2007}•.

@@

Despite the consistent recovery of initial diversity and composition
there was a time lag between the trajectories followed by taxonomic and
functional characteristics. Taxonomic composition had longer time path
and while communities had recovered their functional characteristics,
their taxonomic composition and evenness remained altered by
disturbance. This delay between functional and taxonomic dynamics was
already observed for grasslands \citep{Tilman1997, Mouillot2011} and
more recently for tropical forests \citep{Lohbeck2015, Guariguata2001}.
The functional diversity of communities rely on the functional strategy
of dominant species, according to the ``vegetation quantity effect''
\citep{Grime1998}. The functional trajectories of communities are then
driven first by the increase of diversity and evenness of dominant
species and then by the recruitment of species resembling the old
pre-disturbance community which reduced the overall diversity. Although
dominant species restoredthe functional pre-disturbance characteristics,
communities composition remained altered and the infrequent species of
initial communities remain missing to achieve plots recovery. According
to the funcitonal traits trajectories, those infrequent species would
match the dominant functional type and would participate to the
functional redundancy. The functional redundancy, thas is the functional
overlap between species that is typical of the huge biodiversity of
tropical forests \citep{Bellwood2006}, therefore was not restored 30
years after disturbance and this is majr to consider as it defines
forests' resilience \citep{Trenbath1999, Elmqvist2003, Diaz2005}.

Besides the long-term alteration of functional redundancy, the
persistent compositional changes which might have favored disturbance
resistant species \citep{Haddad2008}, lianas or epiphytes
\citep{Martin2013} and would probably have come along with environmental
change, like in the soils nutrient cycling and compaction
\citep{Olander2005}. These presistent changes highly question forest's
resilience\citep{Chazdon2003a}. New conditions would not only be longer
lasting but self-maintained as tied to disturbance regime
\citep{Burslem2000}. Specificaly, this would impair species contingent
to undisturbed forests, threatening their maintenance, and run the risk
to loose cornerstone species and trigger unexpected ecological
consequences \citep{Jones1994, Diaz2005, Gardner2007}.

@@

Our study showed the significant impact of disturbance on tropical
forests communities. It validated the consistency of the IDH in the long
term and revealed the contrasting response of taxonomic and functional
characteristics with the decoupling between communities taxonomic
evenness and their functional diversity and th dominant functional
strategies. Communities recovery remained unachieved but consistent
recovery of communities assembly for the lowest disturbance intensity
but questioned it after higher disturbance.

>>>>

\caption{\label{tab:Tab1}Intervention table, summary of the disturbance intensity for the 4 plot treatments in Paracou.}
\centering
\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
\toprule
Treatment & Timber & Thinning & Fuelwood & \%AGB lost\\
\midrule
Control &  &  &  & 0\\
T1 & DBH $\geq$ 50 cm, commercial species, $\approx$ 10 trees/ha &  &  & $[12\%-33\%]$\\
T2 & DBH $\geq$ 50 cm, commercial species, $\approx$ 10 trees/ha & DBH $\geq$ 40 cm, non-valuable species, $\approx$ 30 trees/ha &  & $[33\%-56\%]$\\
T3 & DBH $\geq$ 50 cm, commercial species, $\approx$ 10 trees/ha & DBH $\geq$ 50 cm, non-valuable species, $\approx$ 15 trees/ha & 40 cm $\leq$ DBH $\leq$ 50 cm, non-valuable species, $\approx$ 15 trees/ha & $[35\%-66\%]$\\
\bottomrule
\end{tabu}
\end{table}

@@

To counter the remaining taxonomic uncertainty, plots taxonomic
composition and diversity were analysed at the genus level, \emph{i.e.}
referring to the genus of observed or trialed botanical names.
Variations in plots taxonomic and functional composition after
disturbance was visualized by their trajectories in a two-dimensional
ordination space over 30 years. Two NMDS were conducted, either from
taxonomic flora inventories or from plots functional composition based
on the 8 leaf, stem and life history traits. Plots trajectories along
time was reported comparatively to the inventories in 1989, 5 years
after disturbance, which corresponded to first inventory with a
sufficient degree of uncertainty (\textless{}30\% of undetermined
trees). The inventories dissimilarity compared to the reference 1989
inventory was reported using occurrence-based (Jaccard) and
abundance-based (Bray-Curtis) similarity measures. The trajectory of
inventories along time was visualized with the euclidean distance in the
two-dimensional ordination space to the 1989 inventory. The trajectories
of the leaf and stem and life traits were also visualized with the
community weighted means (CWM), representing the average trait value in
a community weighted by relative abundance of the species carrying each
value \citep{Diaz2007, Garnier2004}. To compensate the intrinsinc
difference among plots the trajectories corresponded to the differences
along time with the reference inventory in 1989. Species seed mass
corresponded to 5 classes of increasing mass, seed mass trajectories
were therefore reported as the proportion of each class in the
inventories.

@@

Over time, 828388 trees and 591 botanical species pertaining to 223
genus and 64 botanical families were recorded. Taxonomic and functional
plots trajectories were examined through the trajecotries in a two
dimensional ordination space of successive inventories from 1989 (5
years after disturbance) to 2015 (31 years after disturbance).
Classifications were performed using either abundance-based Bray-Curtis
(Figure \ref{fig:NMDSplans}) or incidence-based Jaccard dissimilarity,
this last having given similar results (data not shown).

@@

The changes observed in plots functional composition went hand to hand
with consistent trajectories of the 8 functional and life history traits
(Leaf thickness, chlorophyll content, toughness and specific area, wood
specific gravity and bark thickness and seed mass and maximum adult
height) visualized with plots community weighted means (CWM) (Figure
\ref{fig:CWM}).

Except for leaf cholorphyll content, which continued to increase for
some T3 and T2 plots 30 years after disturbance, all traits and seed
mass proportions displayed a unimodal trajectories. The weighted means
of communities specific maximum height at adult stage (\emph{Hmax}),
leaf toughness (\emph{L\_toughness}) and wood specific gravity
(\emph{WD}) remained significantly lower than their initial value and
than these of the control plots (Figure \ref{fig:CWM}). The weighted
means of bark thickness (\emph{Bark\_thick}) similarly remained
substantilly higher than initially for all disturbed plots while the
specfic leaf area (\emph{SLA}) had almost recovered its initial value
and this of the undisturbed plots at the end of the experiment. Once
again the maximum difference to initial state was correlated to the
disturbance intensity (\(\rho_{spearman}^{L_{thickness}}=0.67\),
\(\rho_{spearman}^{L_{chloro}}=0.45\),
\(\rho_{spearman}^{L_{toughness}}=-0.43\),
\(\rho_{spearman}^{SLA}=0.93\), \(\rho_{spearman}^{WD}=-0.78\),
\(\rho_{spearman}^{Bark-thickness}=0.88\),
\(\rho_{spearman}^{Hmax}=-0.48\)).

@@

\caption{Trajectories of the communities weighted means (CWM) over 30 years after disturbance of 4 leaf traits (Leaf thickness, \emph{L\_thickness}, chlorophyll content, \emph{L\_chloro}, toughness, \emph{L\_toughness} and specific area, \emph{SLA}), 2 stem traits (wood specific gravity, \emph{WD}, and bark thickness, \emph{Bark-thick}) and one life trait (Specific maximum height at adult stage, \emph{Hmax}). Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation and the missing trait value filling processes. Initial treatments are represented by solid lines colors with green for control, blue for T1,orange for T2 and red for T3.}\label{fig:CWM}
\end{figure*}

@@

For undisturbed plots the Richness, Shannon and Simpson diversities
remained stable over the 30 monitored years. After disturbance the
richness increased when the disturbance intensity was low, with a
maximum increase of 14 botanical genera (plot 3 from treatment 2). When
the disturbance was intense however, plots taxonomic richness followed a
unimodal decrease with a return to initial values after intense
disturbance. The taxonomic evenness (Shannon and Simpson diversities)
significantly increased irrespective of the disturbance intensity. The
evenness followed a unimodal trajectory with a just beginning return
towards initial values and a maximum reached around 20 years and
positively correlated to the disturbance intensity
(\(\rho_{spearman}^{Shannon}=0.86\), and
\(\rho_{spearman}^{Simpson}=0.89\)). Only two T3 plots, plots 8 and 12,
remained increasing 30 years after disturbance \ref{fig:DivTaxo}),
suggesting a similar be much delayed unimodal trajectory.

@@

The functional diversity trajectories of plots based on the 8 leaf, stem
and life history traits was examined through the Rao diversity. For all
undisturbed plots the functional diversity remained comparable to the
reference values 5 years after disturbance (1989 inventories). For all
disturbed plots the functional trajectories followed a unimodal
trajectory with a maximum positively correlated to disturbance intensity
(\(\rho_{spearman}=0.73\)) (Figure \ref{fig:DivFun}). The plot 7 from
treatment 1 that displayed a constantly outlying diversity was removed
from the graphical representation for better readability, see appendix
for full graph. Thirty years after disturbance all plots, whenever the
initial disturbance intensity, regained diversity values similar to
their initial value and to those of control plots.

\textbackslash{}begin\{figure\}

\{\centering \includegraphics[width=1\linewidth]{WholePlotTrajectories_files/figure-latex/DivFun-1}

\}

\textbackslash{}caption\{Upper panels, Trajectories of the Simpson
taxonomic diversity and Rao functional diversity over 30 years after
disturbance. Trajectories correspond to the median (solid line) and
0.025 and 0.975 percentile (gray envelope) observed after 50 iteration
of the taxonomic uncertainty propagation and the missing trait value
filling processes. Initial treatments are represented by solid lines
colors with green for control, blue for T1,orange for T2 and red for T3.
Lower panels, Relationship between the initial \%AGB removed and the
values of Simpson (c) and Rao (d) diversities at three times after
disturbance. Solid lines colors represent the time, 10 years (yellow),
20 years (orange) and 30 years (brown) after
disturbance.\}\label{fig:DivFun} \textbackslash{}end\{figure\}

@@

{\centering \includegraphics[width=1\linewidth]{WholePlotTrajectories_files/figure-latex/RedFun-1}

@@

\caption{Trajectories of the functional redundancy over 30 years after disturbance. Trajectories correspond to the median (solid line) and 0.025 and 0.975 percentile (gray envelope) observed after 50 iteration of the taxonomic uncertainty propagation and the missing trait value filling processes.Initial treatments are represented by solid lines colors with green for control, blue for T1,orange for T2 and red for T3.}\label{fig:RedFun}
\end{figure}

@@

The monitoring of disturbed forest communities confirmed a limited
negative impact of disturbance on species richness, as observed on
several post logging surveys \citep{Cannon1998, Baraloto2012a}. Thirty
years after disturbance, the genus richness was restored to initial plot
values after high disturbance intensity while it substantially increased
after low disturbance intensity, reaching a gain of almost 12 genera for
some plots.

Both richness and evenness followed asymptotic trajectories after
disturbance, sharply increasing for 15 years before stabilizing at
higher values than those of control plots. An increase of communities
evenness stems from a higher homogeneity of species distribution. After
disturbance, communities are made of the old, pre-disturbance survivors
and the newly recruited trees: changes in composition and abundance
distribution are to be found in the recruitment processes and in the
pre-disturbance survivors mortality. Because the composition of old
survivors proved to mirror the initial community \citep{Herault2018}, a
specific turnover was expected among recruited trees with enhanced
growth and survival of previously infrequent species. Indeed, the
increase in taxonomic diversity was accompanied by an increase of
taxonomic dissimilarity with plots initial state and a functional shift
towards resource-acquisitive strategies (sharp increase in the SLA, leaf
thickness and bark thickness and decrease in wood density, leaf
toughness and maximum height)
\citep{Westoby1998, Wright2004, Reich2014}. Disturbance then causes a
reorganization of the typical high dominance structure of hyperdiverse
mature forests after disturbance, benefiting to pioneers and light
demanding species. Likely, the changes in abiotic environment and
competitive pressure favored pioneers which outcompete others in non
limiting resources but are excluded in mature forests by long-lived,
more resistant and shade tolerant species.

As stated by the IDH, communities dynamics after disturbance relied on
species functional strategy and corresponding ability to fill the
environmental niches made available by disturbance. Recruited species
then mixed with pre-disturbance ones, from which they differed, and
constituted a community all the more diversified that the disturbance
was intense \citep{Molino2001}.

@@

All diversity and composition trajectories followed similar hump-back
trajectories returning towards their initial values. Communities
trajectories therefore illustrated their taxonomic and function
resilience.

Because functional traits and functional diversity are the most direct
link between biodiversity and ecosystem functioning \citep{Diaz2005},
their resilience meant a consistent recovery of ecosystem processes in
the long term \citep{Guariguata2001}. The resilience of the taxonomic
diversity and composition similarly translated the recovery of
communities such as before disturbance. This resilience meant the
maintenance of communities initial differences in composition and
supported the idea of their convergence towards a determined state
\citep{Hubbell1999, Molino2001, Baraloto2012a}. If all plots followed
similar hump-back trajectories, though, some significant differences
emerged. To name but one, the taxonomic evenness of one plot displayed
an extensive increase until 30 years after disturbance. The differences
trajectories maxima and time path, along with the fact intrinsinc
communities differences are maintained, suggested that the trajectories
depend on the initial plots composition or on some abiotic parameters
\citep{Anderson2007}.

@@

Despite the recovery of pre-disturbance state there was a time lag
between the trajectories followed by communities taxonomic and
functional characteristics. The trajectory of taxonomic composition and
evenness had a longer time-path and remained altered while communities
functional composition and diversity had already recovered.

Such delay between functional and taxonomic dynamics was already
observed for grasslands \citep{Tilman1997, Mouillot2011} and more
recently for tropical forests \citep{Lohbeck2015, Guariguata2001}.
According to the ``vegetation quantity effect'' \citep{Grime1998}, the
functional diversity of communities rely on the dominant species.
Communities functional trajectories were then driven first by the
increase in diversity and evenness of dominant species, which restored
the pre-disturbance functional characteristics. Then, the functional
trajectories were driven by the recruitment of species resembling the
old pre-disturbance community, which reduced the overall diversity. At
that time, the taxonomic composition and diversity remained altered: the
still missing species were then the initially infrequent and
functionally redundant ones. The functional redundancy, the functional
overlap between species that is typical of the huge biodiversity of
tropical forests \citep{Bellwood2006}, was then not restored 30 years
after disturbance and this is major to consider as it defines forests'
resilience \citep{Trenbath1999, Elmqvist2003, Diaz2005}.

Besides the long-term alteration of functional redundancy, there was
probably persistent compositional changes favoring disturbance resistant
species \citep{Haddad2008}, lianas or epiphytes \citep{Martin2013} and
environmental changes, like in the soils nutrient cycling and compaction
\citep{Olander2005}. These presistent changes highly question forest's
resilience \citep{Chazdon2003a}. New conditions would not only be longer
lasting but self-maintained as tied to disturbance regime
\citep{Burslem2000}. Specificaly, this would impair species contingent
to undisturbed forests, threatening their maintenance, and run the risk
to loose cornerstone species and trigger unexpected ecological
consequences \citep{Jones1994, Diaz2005, Gardner2007}.

@@

Our study showed the significant impact of disturbance on tropical
forests communities and validated the consistency of the IDH in the long
term. It revealed the contrasting response of communities taxonomic and
functional characteristics, with persisting impacts on the species
abundance distibution while the functional diversity and dominant
functional strategies were restored. Communities recovery therefore
remained unachieved but consistent for the range of disturbance studied
here. The length of the recovery, however, severly questioned the
sustainability of intense selective logging and advocated felling cycle
much longer than 30 years \citep{Gourlet-Fleury2005}.
